---
title: "Gapminder Data Analysis"
author: "Isaac Yeoh"
date: "2023-02-19"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## 1. Load and check the data (5 points)

Your first task is to do a very simple data check:

2. (2pt) Load data. How many rows/columns do we have?
``` {r, message=FALSE}
library(tidyverse)
gapminder <- read_delim("data/gapminder.csv")
```
+ We have `r nrow(gapminder)` number of rows and `r ncol(gapminder)` number of columns.

3. (2pt) Print a small sample of data. Does it look OK?
``` {r, include = TRUE}
head(gapminder, 3)
```
+ Yes it looks ok.

## 2. Descriptive statistics (15pt)
1. (3pt) How many countries are there in the dataset? Analyze all three: iso3, iso2 and name.

+ There are `r length(unique(gapminder$iso3))` number of iso3 countries, `r length(unique(gapminder$iso2))` number of iso2 countries, and `r length(unique(gapminder$name))` number of country names

2.

(a) (5pt) Find how many names are there for each iso-2 code. Are there any iso-2 codes that
correspond to more than one name? What are these countries?
```{r, include = TRUE}
gapminder %>%
  group_by(iso2) %>%
  summarize(num = n_distinct(name)) %>%
  arrange(desc(num)) %>%
  head(3)
```
``` {r, include = TRUE}
gapminder %>%
  group_by(iso2) %>%
  filter(is.na(iso2)) %>%
  reframe(country = unique(name))
```
+ As observed in the table, NA has 2 values. The reason for that is because Namibia has
the country code "NA" and there is one other country with no recorded name and iso2 value.

(b) (5pt) Now repeat the same for name and iso3-code. Are there country names that have
more than one iso3-code? What are these countries?
Hint: two of these entitites are CHANISL and NLD CURACAO.

``` {r, include = TRUE}
gapminder %>%
  group_by(name) %>%
  filter(is.na(name)) %>%
  reframe(iso3 = unique(iso3))
```
+ The table above displays the countries with a country iso3 code, but with no name

3. (2pt) What is the minimum and maximum year in these data?
``` {r, include = TRUE}
gapminder %>%
  filter(!is.na(time)) %>%
  summarize(min(time), max(time))
```
## 3. CO2 emissions (30pt)

1. (2pt) How many missing co2 emissions are there for each year? Analyze both missing CO2
and co2_PC. Which years have most missing data?
```{r}
co2_emissions <- gapminder %>%
  filter(is.na(co2)) %>%
  group_by(time) %>%
  summarize(missing_co2 = n()) %>%
  arrange(desc(missing_co2)) %>%
  head(5)
co2_emissions
```
```{r}
co2_pc_emissions <- gapminder %>%
  filter(is.na(co2_PC)) %>%
  group_by(time) %>%
  summarize(missing_co2_PC = n()) %>%
  arrange(desc(missing_co2_PC)) %>%
  head(5)
co2_pc_emissions
```
+ In both tables, years 2017, 2018, and 2019 have the most missing co2 emissions data.

2. (5pt) Make a plot of total CO2 emissions over time for the U.S, China, and India. Add a few more countries of your choice. Explain what do you see.
```{r}
co2_countries <- gapminder %>%
  filter(!is.na(co2), name%in%c("United States of America", "China", "India", "Malaysia", "Poland")) %>%
  group_by(time, name) %>%
  reframe(total_co2_per_year = sum(co2), time) %>%
  arrange(name)
```

```{r}
ggplot(co2_countries) +
  geom_point(aes(x = time, y= total_co2_per_year, col = name)) +
  labs(x = "Time", y = "Total CO2 emissions (kt)")
```

+ We observe China's CO2 emissions increasing exponentially every year. Malaysia's CO2 emissions are also increasing, but not as fast as the rate of India's CO2 emissions. On the other hand, the US's CO2 increased from 1960 to around 2008, but started decrease after that. Poland's CO2 emissions barely has any change across the years.

3. (5pt) Now let’s analyze the CO2 emissions per capita (co2_PC). Make a similar plot of the
same countries. What does this figure suggest?
```{r}
co2_PC_countries <- gapminder %>%
  filter(!is.na(co2_PC), name%in%c("United States of America", "China", "India", "Malaysia", "Poland")) %>%
  group_by(time, name) %>%
  reframe(total_co2_per_year = sum(co2_PC), time) %>%
  arrange(name)
```

```{r}
ggplot(co2_PC_countries) +
  geom_point(aes(x = time, y= total_co2_per_year, col = name)) +
  labs(x = "Time", y = "Total CO2 per capita emissions (kt)")
```

+ The CO2 emission per capita has been increasing for countries like Malaysia and China over time. India's CO2 emission per capita has also been increasing, but its rate is not as significant as Malaysia's and China's. On the other hand, both the US and Poland's CO2 emission per capita have increased from 1960 to 1980, but since then, has been decreasing.

4. (6pt) Compute average CO2 emissions per capita across the continents (assume region is the same as continent). Comment what do you see. Note: just compute averages over countries and ignore the fact that countries are of different size. Hint: Americas 2016 should be 4.80.

``` {r}
region_CO2 <- gapminder %>%
  filter(!is.na(region), !is.na(co2_PC)) %>%
  group_by(region, time) %>%
  reframe(av_CO2_PC = mean(co2_PC))
```

```{r}
ggplot(region_CO2) +
  geom_point(aes(x = time, y= av_CO2_PC, col = region)) +
  labs(x = "Time", y = "Average CO2 per capita emissions (kt)")
```

+ Africa has had a steady increase in average CO2 per capita over time. Every other region had varying average CO2 per capita over time between the 1960s to 1980s. From there, we can observe Europe's average CO2 per capita over time decreasing, but Asia, Oceania, and the Americas increasing.

5. (7pt) Make a barplot where you show the previous results–average CO2 emissions per capita
across continents in 1960 and 2016.

```{r}
years <- c("1960", "2016")
region_CO2 %>%
  filter(time%in%years) %>%
  ggplot(aes(x=region, y=av_CO2_PC, fill = factor(time))) +
  geom_col(position = "dodge") +
  labs(x = "Region", y = "Average CO2 per capita", fill = "Year")
```

6. Which countries are the three largest, and three smallest CO2 emitters (in terms of CO2 per capita) in 2019 for each continent? (Assume region is continent).

+ Since there is no CO2 per capita data in 2019, I used the 2016 data.

+ Smallest CO2 emitters in 2016 for each continent:
```{r}
gapminder %>%
  filter(!is.na(region), !is.na(co2_PC), time == "2016") %>%
  group_by(region, name) %>%
  summarize(smallest_CO2_emitters_2016 = mean(co2_PC), .groups = "drop") %>%
  arrange(region, smallest_CO2_emitters_2016) %>%
  group_by(region) %>%
  slice(1:3)
```

+ Largest CO2 emitters in 2016 for each continent:
```{r}
gapminder %>%
  filter(!is.na(region), !is.na(co2_PC), time == "2016") %>%
  group_by(region, name) %>%
  summarize(largest_CO2_emitters_2016 = mean(co2_PC), .groups = "drop") %>%
  arrange(region, desc(largest_CO2_emitters_2016)) %>%
  group_by(region) %>%
  slice(1:3)
```

## GDP per capita (50pt)
1. (8pt) Make a scatterplot of GDP per capita versus life expectancy by country, using data for 1960. Make the point size dependent on the country size, and color those according to the continent. Feel free to adjust the plot in other ways to make it better. Comment what do you see there.

```{r}
gapminder %>%
  filter(!is.na(GDP_PC), !is.na(lifeExpectancy), !is.na(region), time == "1960") %>%
  group_by(name) %>%
  ggplot(aes(x = GDP_PC, y = lifeExpectancy, col = region)) +
  geom_point(aes(size = totalPopulation)) +
  labs(x = "GDP per Capita", y = "Life Expectancy", size = "Total Population", col = "Region")
```

2. (4pt) Make a similar plot, but this time use 2019 data only.
```{r}
gapminder %>%
  filter(!is.na(GDP_PC), !is.na(lifeExpectancy), !is.na(region), time == "2019") %>%
  group_by(name) %>%
  ggplot(aes(x = GDP_PC, y = lifeExpectancy, col = region)) +
  geom_point(aes(size = totalPopulation)) +
  labs(x = "GDP per Capita", y = "Life Expectancy", size = "Total Population per country", col = "Region")
```

3. (6pt) Compare these two plots and comment what do you see. How has world developed
through the last 60 years?
+ In the last 60 years, the total population per country has increased as seen by the size of the points. Alongside that, life expectancy has increased, and GDP per capita has significantly increased as well.

4. (6pt) Compute the average life expectancy for each continent in 1960 and 2019. Do the results fit with what do you see on the figures? Note: here as average I mean just average over countries, ignore the fact that countries are of different size.

```{r} 
gapminder %>%
  filter(!is.na(time), time%in%c("1960", "2019"), !is.na(lifeExpectancy)) %>%
  group_by(time) %>%
  summarize(mean(lifeExpectancy))
```

+ Yes the results match what I see on the figures

5. (8pt) Compute the average LE growth from 1960-2019 across the continents. Show the results in the order of growth. Explain what do you see.
Hint: these data (data in long form) is not the simplest to compute growth. But you may
want to check out the lag() function. And do not forget to group data by continent when
using lag(), otherwise your results will be messed up

```{r}
gapminder %>%
  filter(!is.na(region), !is.na(lifeExpectancy), time%in%c(1960, 2019)) %>%
  group_by(region, time) %>%
  summarize(avg_LE = mean(lifeExpectancy)) %>%
  mutate(previous = lag(avg_LE), growth = avg_LE - previous) %>%
  filter(!is.na(growth)) %>%
  arrange(desc(growth))
```

6. (6pt) Show the histogram of GDP per capita for years of 1960 and 2019. Try to put both
histograms on the same graph, see how well you can do it!
```{r}
gapminder %>%
  filter(!is.na(GDP_PC), time%in%c("1960", "2019")) %>%
  ggplot(aes(x = GDP_PC, fill = factor(time))) +
  geom_histogram(position = "dodge") +
  labs(x = "GDP per capita between 1960 and 2019", fill = "Year")
```

7. (6pt) What was the ranking of US in terms of life expectancy in 1960 and in 2019? (When
counting from top.)
```{r}
gapminder %>%
  filter(!is.na(name), !is.name(lifeExpectancy), time%in%c(1960, 2019)) %>%
  group_by(time) %>%
  mutate(rank = rank(desc(lifeExpectancy))) %>% 
  filter(name == "United States of America") %>%
  summarize(time, rank)
```

8. (6pt) If you did this correctly, then you noticed that US ranking has been falling quite a bit. But we also have more countries in 2019–what about the relative rank divided by the
corresponding number of countries that have LE data in the corresponding year?
```{r}
gapminder %>% 
    filter(time%in%c(1960, 2019), !is.na(lifeExpectancy), !is.na(region)) %>% 
    group_by(time) %>%
    mutate(le_rank = rank(desc(lifeExpectancy)), num_country = n(), relative_rank = le_rank / num_country) %>% 
    select(name, le_rank, time, num_country, relative_rank) %>% 
    filter(name == "United States of America")
```

# I spent 6 hours on this assignment
